Microxanox: an R package for simulating an aquatic \(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.

### 2022-06-16 inst/manuscript/manuscript_microxanox.tex
Changes Bibliography and re-compiled.

  \begin{abstract}
  \(Microxanox\) is an R package to simulate a three functional group
  system (cyanobacteria, phototrophic sulfur bacteria, and
  sulfate-reducing bacteria) with four chemical substrates (phosphorus,
  oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary
  differential equations. Simulations can be run individually or over a
  parameter range, to find stable states. The model features multiple
  species per functional group, where the number is only limited by
  computational constraints. The R package is constructed in such a way,
  that the results contain the input parameter used, so that a saved
  results can be loaded again and the simulation be repeated.
  \end{abstract}
    \begin{keyword}
    reproducibility \sep regime shift \sep stable state \sep 
    ordinary differential equations
  \end{keyword}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule()
\endhead
C1 & Current code version & v0.9.0 \\
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
& & \href{https://zenodo.org/record/6624125}{10.5281/zenodo.6624125} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & MIT \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
C8 & If available Link to developer documentation/manual &
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{Usert
Manual} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Owen.Petchey@ieu.uzh \\
\bottomrule()
\end{longtable}

@@

Mathematical models play a key role in the development of understanding
about how ecosystems work and how they respond to environmental changes
\citep{Binzer2016a, Chaparro-Pedraza2021, Vasseur2005}. They are also
important for developing hypotheses that can then be tested in empirical
studies. One area of ecology in which simple models have played a
influential role is the area of how ecosystems respond to gradual change
in an environmental driver \citep{Scheffer2001}. An environmental driver
is a environmental condition that affects an ecosystem, but is assumed
to not be affected by the ecosystem, such as the rate of nutrient input
into a lake.

@@

An example where a gradual change of an environmental variable causes an
abrupt change of the system is the switch from an aerobic (oxygen is
available for metabolism) to anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. Three types of microbes occur in the
model: cyanobacteria (CB) dominating the oxic state, and two types of
sulfur bacteria that dominate the anoxic state (sulfate reducing
bacteria (SB) and phototrophic sulfur bacteria (PB)). This system has
been investigated by \citet{Bush2017} in a simulation study of a
mathematical model. They showed that gradual change in the rate at which
oxygen could diffuse into the ecosystem (termed the oxygen
diffusitivity) could cause catastrophic changes in the ecosystem state
that would be difficult to reverse.

One feature of the study by \citet{Bush2017} was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves open the question of if and how biodiversity
within these types (i.e.~functional groups) of bacteria, in their model,
affects the ecosystem response to environmental change. This limitation
is not specific to the study of \citet{Bush2017}, there are few if any
studies of the effects of biodiversity on abrupt transitions between
ecosystem states.

@@

The second stage was to add functionality that would be necessary to
answer our research question. Most importantly, we made it possible to
have multiple species of bacteria within each of the three functional
groups, for the multiple species to differ in their characteristics, and
to vary the number of species and amount of variability among them. We
also added functionality that allowed: temporally varying environmental
conditions, addition of random noise to state variables, and
immigration. In addition to the model itself, the package includes some
functions to analyse the results as well as visualize the results to
provide a starting point for customized visualizations based on own
requirements. The basic and additional functionality is described in the
package
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

@@

The framework used when writing this package aimed to maximise
simplicity for the user, and to make it easy to reproduce results (see
the supplement \citep{Petchey2022} to \citet{Limberger2022} for an
example of how this is used). As such, all the parameters needed to run
a simulation or find a stable state (i.e.~the final state of the
ecosystem) are contained in a single object (which can easily be created
using included functions). This parameter object is given to a function
that runs the simulations and returns the results. The returned results
object is identical to the parameter object but with an additional slot
named \texttt{results}, which contains the results of the run. Thus the
returned results object contains the simulation conditions (parameters)
as well as the results, and can be used to run the simulation again.
This promotes reproducibility and makes incremental changes of
individual parameters and re-running the simulations straightforward.

@@

Two methods for finding steady states are implemented. The first runs a
separate simulation for each combination of starting conditions and
oxygen diffusivity (we term this the \emph{Replication method}). This is
the method used in the \citet{Bush2017} study. The second runs two
simulations, with step-wise and slowly temporally increasing oxygen
diffusivity, and the other with step-wise and slowly decreasing or
decreasing oxygen diffusivity. During this temporal environmental
change, the state of the system was recorded just before change to a new
oxygen diffusivity (we term this the \emph{Temporal method}). We
implemented two methods since there is no definitive best method, and in
order to check if results were sensitive to choice of method.

The replication method is implemented in the function
\texttt{run\_replication\_ssfind()} which takes a parameter object as
returned by the function \texttt{new\_replication\_ssfind\_parameter()}
and the number of cores for multithreading the simulation. As the
multithreading uses the package function \texttt{mclapply()} from the
package \texttt{parallel} \citep{RCoreTeam2022}, the multthreading only
works on Linux and Mac. It is planned to move to \texttt{parLapply()}
\citep{RCoreTeam2022} in a future release.

This temporal method implemented in the function
\texttt{run\_temporal\_ssfind()}, which takes a parameter object as
created by the function \texttt{new\_temporal\_ssfind\_parameter()}. It
is planned for a later release, to run these two simulations in
parallel.

@@

{\centering \includegraphics[width=350px]{/Users/rainer/git/0.RPackages/microxanox/inst/manuscript/figures/gen_uc1_partrep_temporal_state_switching}

@@

Using the package to calculate the extent of hysteresis involves setting
ecosystem and simulation parameters, including parameters for the
finding of stable states across an environmental gradient, running the
stable state finding function, and analysing the results with the
function that calculates extent of hysteresis. The code for this is
provided in the \emph{User Guide} vignette.

@@

{\centering \includegraphics[width=150px]{/Users/rainer/git/0.RPackages/microxanox/inst/manuscript/figures/gen_uc2_user_guide_hysteresis}

>>>>

  \begin{abstract}
  \(Microxanox\) is an R package to simulate a three functional group
  ecosystem (cyanobacteria, phototrophic sulfur bacteria, and
  sulfate-reducing bacteria) with four chemical substrates (phosphorus,
  oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary
  differential equations. Simulations can be run individually or over a
  parameter range, to find stable states. The model can be implemented
  with different numbers of species per functional group. The package is
  constructed in such a way that the results contain the input parameter
  used, so that a saved results can be loaded again and the simulation
  be repeated. Furthermore, the package framework and code should serve
  as a useful starting point for making simulation models of other types
  of ecosystem.
  \end{abstract}
    \begin{keyword}
    reproducibility \sep regime shift \sep stable state \sep 
    ordinary differential equations
  \end{keyword}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule
\endhead
C1 & Current code version & v0.9.0 \\
C2 & Permanent link to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
& & \href{https://zenodo.org/record/6624125}{10.5281/zenodo.6624125} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & MIT \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
C8 & If available Link to developer documentation/manual &
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Owen.Petchey@ieu.uzh \\
\bottomrule
\end{longtable}

@@

Mathematical models play a key role in the development of understanding
about how ecosystems work and how they respond to environmental changes
\citep{Binzer2016a, Chaparro-Pedraza2021, Vasseur2005}. They are also
important for developing hypotheses to test in empirical studies. One
area of ecology in which models have played a influential role is how
ecosystems respond to gradual change in an environmental driver
\citep{Scheffer2001}. An environmental driver is a environmental
condition that affects an ecosystem, but is assumed to not be affected
by the ecosystem, such as the rate of nutrient input into a lake.

@@

An example where a gradual change of an environmental variable causes an
abrupt change of the system is the switch from an aerobic (oxygen is
available for metabolism) to anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. This system has been investigated by
\citet{Bush2017} in a simulation study of a mathematical model. Three
types of microbes occur in the model: cyanobacteria (CB) dominating the
oxic state, and two types of sulfur bacteria that dominate the anoxic
state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria
(PB)). The model shows that gradual change in the rate at which oxygen
could diffuse into the ecosystem (termed the oxygen diffusitivity) could
cause catastrophic changes in the ecosystem state that would be
difficult to reverse.

One feature of the study by \citet{Bush2017} was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves open the question of if and how biodiversity
within these types (i.e.~functional groups) of bacteria affects the
ecosystem response to environmental change. This limitation is not
specific to the study of \citet{Bush2017}. There are few if any studies
of the effects of biodiversity on abrupt transitions between ecosystem
states.

@@

The second stage was to add functionality that would be necessary to
answer our research question. Most importantly, we made it possible to
have multiple species of bacteria within each of the three functional
groups, for the multiple species to differ in their characteristics, and
to vary the number of species and amount of variability among them. We
also added functionality that allowed: temporally varying environmental
conditions, addition of random noise to state variables, and
immigration. In addition to the model itself, the package includes some
functions to analyse the results as well as to visualize the results to
provide a starting point for customized visualizations based on own
requirements. The basic and additional functionality is described in the
package
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

@@

The framework used when writing this package aimed to maximise
simplicity for the user, and to make it straightforward to reproduce
results (see the supplement \citep{Petchey2022} to \citet{Limberger2022}
for an example of how this is used). As such, all the parameters needed
to run a simulation or find a stable state (i.e.~the final state of the
ecosystem) are contained in a single object (which can easily be created
using included functions). This parameter object is given to a function
that runs the simulations and returns the results. The returned results
object is identical to the parameter object but with an additional slot
named \texttt{results}, which contains the results of the run. Thus the
returned results object contains the simulation conditions (parameters)
as well as the results, and can be used to run the simulation again.
This promotes reproducibility and makes incremental changes of
individual parameters and re-running the simulations straightforward.

@@

Two methods for finding steady states are implemented. The first runs a
separate simulation for each combination of starting conditions and
oxygen diffusivity (we term this the \emph{Replication method}). This is
the method used in the \citet{Bush2017} study. The second runs two
simulations, one with step-wise and slowly temporally \emph{increasing}
oxygen diffusivity, and the other with step-wise and slowly
\emph{decreasing} oxygen diffusivity. During this temporal environmental
change, the state of the system is recorded just before change to a new
oxygen diffusivity (we term this the \emph{Temporal method}). We
implemented two methods since there is no definitive best method, and in
order to check if results were sensitive to choice of method.

The replication method is implemented in the function
\texttt{run\_replication\_ssfind()} which takes a parameter object as
returned by the function \texttt{new\_replication\_ssfind\_parameter()}
and the number of cores for multithreading the simulation. As the
multithreading uses the package function \texttt{mclapply()} from the
package \texttt{parallel} \citep{RCoreTeam2022}, the multithreading only
works on Linux and Mac. It is planned to move to \texttt{parLapply()}
\citep{RCoreTeam2022} in a future release.

The temporal method is implemented in the function
\texttt{run\_temporal\_ssfind()}, which takes a parameter object as
created by the function \texttt{new\_temporal\_ssfind\_parameter()}. It
is planned for a later release, to run these two simulations in
parallel.

@@

{\centering \includegraphics[width=350px]{figures/gen_uc1_partrep_temporal_state_switching}

@@

Using the package to calculate the extent of hysteresis involves setting
ecosystem and simulation parameters, including parameters for the
finding of stable states across an environmental gradient, running the
stable state finding function, and analysing the results with the
function that calculates extent of hysteresis. The code for this is
provided in the \emph{User Guide}.

@@

{\centering \includegraphics[width=150px]{figures/gen_uc2_user_guide_hysteresis}
